---
title: "Correlating like a Data Master"
author: "VISHAL BHASHYAAM"
format: html
editor: visual
---

# Correlating Like a Data Master

## Purpose of this chapter

Assess relationships within a novel data set using publication quality tables and plots

## Take-aways

1.  Describe and visualize correlations between numerical variables

2.  Visualize correlations of all numerical variables within groups

3.  Describe and visualize relationships based on target variables

### Required setup

We first need to prepare our environment with the necessary packages.

```{r}
if (!require(pacman))
  install.packages("pacman")

pacman::p_load(
       dlookr,
       formattable,
       GGally,
       ggdist,
       ggpubr,
       ggridges,
       here,
       tidyverse)

# Set global ggplot() theme
# Theme pub_clean() from the ggpubr package with base text size = 16
theme_set(theme_pubclean(base_size = 12)) 
# All axes titles to their respective far right sides
theme_update(axis.title = element_text(hjust = 1))
# Remove axes ticks
theme_update(axis.ticks = element_blank()) 
# Remove legend key
theme_update(legend.key = element_blank())
```

## **Load the Examine a Data Set**

```{r}
# Loading the data 
dataset <- dataset <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-08-22/population.csv')
populations2010 <- filter(dataset, year >= 2010)
write_csv(populations2010, "population.csv")
```

```{r}
# What the data looks like 

dataset |>
  head()|>
formattable()

dataset |>
  head() |>
  formattable()
```

## Describe and Visualize Correlations

Correlations are a statistical relationship between two numerical variables, may or may not be casual. Exploring Correlations in data allows to determine data independence , assumptions of parametric statistics which means the variables are both randomly collected

```{r}
# Table of correlations between numerical variables
correlate(dataset) |>
  formattable()
```

```{r}
# Correlation matrix of numerical variable 
dataset |>
  plot_correlate()
```

## Visualize Correlations within groups

Let's see how each group co-related within these groups .

```{r}
dataset |>
  group_by(coo) |>
  plot_correlate()
```

This is great, we have our correlations within groups! However, the correlation matrices aren't always the most intuitive, so let's plot!

We will be using the `ggpairs()` function within the `GGally` package. Specifically, we are looking at the correlations between predawn leaf water potential `pLWP` and midday leaf water potential `mLWP`. Leaf water potential is a key indicator for how stressed plants are in droughts.

```{r}
dataset |> 
  dplyr::select(refugees,returned_refugees,asylum_seekers) |>
  ggpairs(aes(color="red" ,alpha = 0.5)) +
  theme(strip.background = element_blank())  # Removing facet strip backgrounds

```

-   Since the dataset - Refugee Dataset I use does not contain all the required data for Operations I am changing the data to the default dataset used in the github.

```{r}
data <- read.csv(".\\Data\\Figure 2\\Data_Fig2_Repo.csv")
data["Sap_Flow"] =data["SF"]
data |>
  head()|>
  formattable()
```

## Describe and Visualize Relationships Based on Target Variables

### Target Variables

`Target variables` are essentially numerical or categorical variables that you want to relate others to in a data frame. `dlookr` does this through the `target_by()` function, which is similar to `group_by()` in `dplyr`. The `relate()` function then briefly analyzes the relationship between the target variable and the variables of interest.

The relationships below will have the formula relationship `target ~ predictor`.

### Numerical Target Variables: Numerical Variable of Interest

`Formula: Sap_Flow (numerical response)  ~ pLWP (numerical predictor)`

```{r}
# First, we need to remove NAs, they cause an error
data.noNA <- data |> 
  drop_na()

# The numerical predictor variable that we want
num <- target_by(data.noNA, Sap_Flow)

# Relating the variable of interest to the numerical target variable
num_num <- relate(num, pLWP)

# Summary of the regression analysis - the same as the summary from lm(Formula)
summary(num_num)


```

```{r}
# Plot the linear relationship
plot(num_num) 

```

### **Numerical Target Variables: Categorical Variable of Interest**

Formula: `pLWP (numerical response) ~ Group (categorical predictor)`

```{r}
# The categorical predictor variable that we want
num <- target_by(data, pLWP) 

# We need to change Group to a factor
num$Group <- as.factor(num$Group)

# Relating the variable of interest to the numerical target variable
num_cat <- relate(num, Group)

# Summary of the ANOVA analysis - the same as the summary from anova(lm(Formula))
summary(num_cat)

```

```{r}
plot(num_cat) + 
  theme(axis.text.x = element_blank())
```

### Categorical Target Variables: Numerical Variable of Interest

Note that this produces descriptive statistics, unlike the other relationships we are looking at.

Formula: `Group (categorical) ~ pLWP (numerical)`

```{r}
# The categorical predictor variable that we want
categ <- target_by(data, Group)

# Relating the variable of interest to the numerical target variable
cat_num <- relate(categ, pLWP)

# Summary of descriptive statistics
summary(cat_num)
```

```{r}
plot(cat_num)
```

## Categorical Target Variables: Categorical Variable of Interest

We only have 1 categorical variable, so creating another.

```{r}
# Create new categorical column
cat_dataset <- data |>
  select(pLWP, Group) |>
  drop_na() |>
  mutate(HighLWP = ifelse(
    pLWP > (mean(pLWP + sd(pLWP))), 
                          "Yes", 
                          "No"))

# New dataset 
cat_dataset |>
  head() |>
  formattable()


```

Now we have two categories!

```{r,error=TRUE}
# The categorical predictor variable that we want
categ <- target_by(cat_dataset, HighLWP)

# Relating the variable of interest to the categorical target variable
cat_cat <- relate(categ, Group)

# Summary of the Chi-square test for Independence
summary(cat_cat)
```

```{r,error=TRUE}
plot(cat_cat)
```
