---
title: "Exploring like a Data Adventure"
author: "VISHAL BHASHYAAM"
format: html
editor: visual
---

# Exploring like a Data Adventure

## Purpose of this chapter

### Exploring the normality of numerical columns in a novel data set and producing publication quality tables and reports

## Take-aways

1.  Using summary statistics to better understand individual columns in data sets.

2.  Assessing data normality in numerical columns.

3.  Producing a publishable HTML with summary statistics and normality tests for columns within a data set.

## Required Setup

Prepare the environment with required packages

```{r}
# Sets the number of significant figures to three -e.g, 1.333
options(digits = 3)

# Downloading required package for quick package loading
if(!require(pacman))
  installed.packages("pacman")
#loading package
pacman::p_load(dlookr, #Exploratory data analysis(EDA)
               formattable, # HTML tables from R outputs
               here, # Standardizes paths to data 
               kableExtra, # Alternative to formattable 
               knitr, # Needed to write HTML reports
               tidyverse # Powerful data wrangling package suite like Pandas for python
               )
```

## ̥Load and Examine the Dataset

We will be using Refugees R package,

This package provides data from three major sources:

-   Data from UNHCR's annual statistical activities dating back to 1951.

-   Data from the United Nations Relief and Works Agency for Palestine Refugees in the Near East (UNRWA), specifically for registered Palestine refugees under UNRWA's mandate.

-   Data from the Internal Displacement Monitoring Centre (IDMC) on people displaced within their country due to conflict or violence.

The {refugees} package includes eight datasets. I will be working with `population` with data from 2010 to 2022.

```{r}
dataset <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-08-22/population.csv')
populations2010 <- filter(dataset, year >= 2010)
write_csv(populations2010, "population.csv")

```

```{r}
# What does the data look like
dataset |> 
  head() |> 
   formattable()

dataset |> 
  tail() |> 
   formattable()
```

## Diagnose your Data

```{r}
# Porperties of Data 
dataset |> 
  diagnose() |>
  formattable()
```

-   `variables` : name of each variable in the column

-   `types` : data type of each variable

-   `missing_count` : number of missing values

-   `missing_percent` : percentage of missing values

-   `unique_count`: number of unique values

-   `unique_rate` : rate of unique values - unique_count / Number of observations

## Box Plot

## Skewness:

### Note:

-   Skewness is designed for distribution with one peak (Unimodal), it is meaningless for distribution with multiple peaks (multimodal).

-   Can't make assumptions/conclusions about the locations of mean and median based on the skewness sign.

## Kurtosis

### Note:

-   Kurtosis may work fine for distribution with one peak (Unimodal), it is meaningless for distribution with multiple peaks ( Multimodal).

-   The classic definition of kurtosis is not robust: it could be easily spoiled by extreme outliers.

## Describe your continuous data

```{r}
# Summary statistics

dataset |>
  describe() |>
  formattable()
```

-   `describes_variables`: name of the column being described

-   `n`: number of observations excluding missing values

-   `na`: number of missing values

-   `mean`: arithmetic average

-   `sd`: standard deviation

-   `se_mean`: standard error mean. sd/sqrt(n)

-   `IQR`: interquartile range (Q3-Q1)

-   `skewness`: skewness

-   `kurtosis`: kurtosis

-   `p25`: Q1. 25% percentile

-   `p50`: median. 50% percentile

-   `p75`: Q3. 75% percentile

-   `p01`, `p05`, `p10`, `p20`, `p30`: 1%, 5%, 20%, 30% percentiles

-   `p40`, `p60`, `p70`, `p80`: 40%, 60%, 70%, 80% percentiles

-   `p90`, `p95`, `p99`, `p100`: 90%, 95%, 99%, 100% percentiles

## Describe the continuous data : Refined

Selecting the required values for summary statistics

```{r}
# summary statistics, selecting the desired values 

dataset  |> 
  describe() |>
  select(described_variables, n, na, mean,sd, se_mean, IQR,skewness,kurtosis,p25,p50,p75) |>
  
  formattable()

```

## Describe categorical variables

```{r}
# diagnosis of categorical variables 

dataset |>
  diagnose_category() |>
  formattable()
```

-   `variables`: category names

-   `levels`: group names within categories

-   `N`: number of observation

-   `freq`: number of observation at group level / number of observation at category level

-   `ratio`: percentage of observation at group level / number of observation at category level

-   `rank`: rank of the occupancy ratio of levels (order in which the groups are in the category)

## Group Descriptive Statistics

```{r}
dataset |>
  group_by(coo) |>
  describe() |>
  select(described_variables, n, na, mean,sd, se_mean, IQR,skewness,kurtosis,p25,p50,p75) |>
  
  formattable()
```

## Testing Normality

-   Shapiro-Wilk test & Q-Q plots

-   Testing overall normality of two columns

-   testing normality of groups

## Normality of columns

### Shapiro-Wilk Test

Shapiro-Wilk test looks at whether a target distribution is sample form a normal distribution

```{r}
dataset |> 
  normality() |>
  formattable()
```

## Q-Q Plots

Plots of the quartiles of a target data set and plot it against predicted quartiles from a noraml distribution

```{r}
dataset |> 
  plot_normality()

```

## Normality within groups

Looking within Coo at the subgroup normality

```{r}
dataset |> 
  group_by(coo) |>
  select(refugees,asylum_seekers,returned_refugees,coo) |>
  normality() |>
  formattable()
```

## Q-Q Plots

```{r}
dataset |> 
  group_by(coo) |>
   select(refugees,asylum_seekers,returned_refugees,coo) |>
  plot_normality() 
 
```

## Producing an HTML Normality Summary

```{r}
eda_web_report(dataset,author = "VISHAL BHASHYAAM",subtitle = "Refugee dataset",output_file = "Diagnose_Report_hw-02-02")
```
